---
title: "Supply and Demand"
subtitle: "Chapter 3"
author: "ECON 1450: Principles of Microeconomics"
date: "Fall 2026"
institute: "University of Vermont"
format:
  beamer:
    theme: Madrid
    colortheme: default
    fonttheme: structurebold
    aspectratio: 169
    navigation: horizontal
    section-titles: false
    toc: false
    slide-level: 2
    incremental: false
    cite-method: natbib
header-includes: |
  \usepackage{tikz}
  \usetikzlibrary{shapes,arrows,positioning}
  \definecolor{TealPrimary}{RGB}{0,150,136}
  \definecolor{DarkGray}{RGB}{66,66,66}
  \definecolor{LightGray}{RGB}{240,240,240}
  \setbeamercolor{structure}{fg=TealPrimary}
  \setbeamercolor{palette primary}{fg=white,bg=TealPrimary}
  \setbeamercolor{frametitle}{fg=DarkGray,bg=white}
  \setbeamercolor{block title}{fg=TealPrimary,bg=LightGray!50}
  \setbeamercolor{itemize item}{fg=TealPrimary}
  \setbeamerfont{title}{size=\huge,series=\bfseries}
  \setbeamerfont{frametitle}{size=\Large,series=\bfseries}
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{footline}[frame number]
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false

# Load packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(grid)
library(gridExtra)

# Set global theme for all graphs
theme_modern_econ <- function(base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = rel(1.1), color = "#424242"),
      plot.subtitle = element_text(hjust = 0.5, size = rel(0.9), color = "#5A5A5A"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#F0F0F0", linewidth = 0.3),
      axis.line = element_line(color = "#424242", linewidth = 0.6),
      axis.text = element_text(size = rel(0.95), color = "#424242"),
      axis.title = element_text(size = rel(1), face = "plain", color = "#424242"),
      axis.ticks = element_line(color = "#424242", linewidth = 0.4),
      axis.ticks.length = unit(0.15, "cm"),
      legend.position = "none",
      plot.margin = margin(10, 10, 10, 10)
    )
}

# Color palette
col_demand <- "#424242"  # Dark gray
col_supply <- "#009688"  # Teal
col_equilibrium <- "#009688"
col_shift <- "#00BCD4"  # Light teal

# Helper function to create supply curves
create_supply <- function(intercept = -0.5, slope = 0.15, Q_range = c(7, 15)) {
  data.frame(Q = seq(Q_range[1], Q_range[2], by = 0.05)) %>%
    mutate(P = intercept + slope * Q)
}

# Helper function to create demand curves
create_demand <- function(intercept = 2.5, slope = -0.15, Q_range = c(7, 15)) {
  data.frame(Q = seq(Q_range[1], Q_range[2], by = 0.05)) %>%
    mutate(P = intercept + slope * Q)
}

# Standard supply and demand curves
demand_std <- create_demand()
supply_std <- create_supply()

# Equilibrium
eq_Q <- 10
eq_P <- 1.0
```

#

\begin{center}
{\Huge\textbf{Supply and Demand}}

\vskip0.5em

{\Large Chapter 3}

\vskip2em

{\large ECON 1450: Principles of Microeconomics}

{\normalsize University of Vermont • Fall 2026}
\end{center}

# Opening: A Market in Action

## How much for a hotel room in Burlington on April 7, 2024?

::: {.columns}
:::: {.column width="58%"}
**Typical spring weekend:**
- Comfort Inn: \$120
- Hilton Garden Inn: \$150

\vskip1em

**Eclipse weekend (April 7-8, 2024):**
- Comfort Inn: \$475
- Hilton Garden Inn: \$860
- Many hotels: fully booked

\vskip1em

::: {.block}
### What changed?
Same hotels, same rooms, same city... dramatically different prices. **Supply and demand in action.**
:::
::::

:::: {.column width="40%"}
\begin{center}
\includegraphics[width=\linewidth,height=0.7\textheight,keepaspectratio]{images/eclipse_hotel.jpg}

{\small \textit{Image: Burlington during 2024 total solar eclipse}}
\end{center}
::::
:::

## By October 2023, the picture was clear

Accommodation listings across Vermont were being booked months in advance:

- La Quinta Inn \& Suites: \$624/night
- Best Western Plus: \$475/night
- Hilton Garden Inn: \$860/night
- VRBO and Airbnb: elevated pricing throughout the region

\vskip1em

::: {.block}
#### Market Response
Hotels didn't build new rooms. They didn't change amenities. They simply adjusted the price to match the surge in demand from eclipse viewers.
:::

# Course Roadmap

## Today's Journey

\begin{center}
\begin{tikzpicture}[node distance=2cm, auto]
  \node[rectangle, draw=TealPrimary, thick, fill=TealPrimary!10, text width=3cm, align=center, minimum height=1cm] (curves) {Supply and Demand Curves};
  \node[rectangle, draw=TealPrimary, thick, fill=TealPrimary!10, text width=3cm, align=center, minimum height=1cm, below of=curves] (shifts) {Shifts vs. Movements};
  \node[rectangle, draw=TealPrimary, thick, fill=TealPrimary!10, text width=3cm, align=center, minimum height=1cm, below of=shifts] (eq) {Equilibrium};
  \node[rectangle, draw=TealPrimary, thick, fill=TealPrimary!10, text width=3cm, align=center, minimum height=1cm, below of=eq] (apply) {Real-World Applications};

  \draw[->, thick, TealPrimary] (curves) -- (shifts);
  \draw[->, thick, TealPrimary] (shifts) -- (eq);
  \draw[->, thick, TealPrimary] (eq) -- (apply);
\end{tikzpicture}
\end{center}

## Learning Objectives

By the end of today's class, you will be able to:

1. **Define** competitive markets and explain how supply and demand describe them

2. **Distinguish** between movements along curves versus shifts of curves

3. **Identify** the five factors that shift demand and five that shift supply

4. **Find** market equilibrium and explain how prices eliminate shortages and surpluses

5. **Predict** how price and quantity change when markets experience shocks

\vskip1em

::: {.block}
#### Core Skill
The supply-demand framework is economics' most powerful analytical tool. Master it, and you can explain everything from hotel prices to labor markets to cryptocurrency.
:::

# Part I: The Demand Side

## Competitive Markets: Our Starting Point

For now, we analyze **competitive markets**:

:::: {.columns}
::: {.column width="48%"}
**Characteristics:**

- Many buyers and sellers
- No single participant can influence market price
- Standardized products
- Free entry and exit
:::

::: {.column width="48%"}
**Examples:**

- Agricultural commodities (wheat, corn)
- Stock markets
- Online marketplaces
- Gasoline stations in a city
:::
::::

\vskip1em

::: {.block}
#### Why start here?
Not all markets are competitive (we'll study monopolies and oligopolies later), but competitive markets provide the clearest insights into how prices coordinate economic activity.
:::

## The Demand Curve

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

# Points ON the demand curve (matching demand_std equation: P = 2.5 - 0.15*Q)
demand_schedule <- data.frame(
  Quantity = c(7, 8, 9, 10, 11, 12, 13),
  Price = c(2.5 - 0.15*7, 2.5 - 0.15*8, 2.5 - 0.15*9, 2.5 - 0.15*10,
            2.5 - 0.15*11, 2.5 - 0.15*12, 2.5 - 0.15*13)
)

ggplot() +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.5) +
  geom_point(data = demand_schedule, aes(x = Quantity, y = Price),
             color = col_demand, size = 3) +
  annotate("text", x = 12.5, y = 0.75, label = "Demand",
           color = col_demand, size = 5, hjust = 0, fontface = "bold") +
  annotate("text", x = 9.5, y = 1.65,
           label = "As price falls,\nquantity rises",
           color = col_demand, size = 3.2, hjust = 0.5) +
  annotate("segment", x = 9.8, y = 1.45, xend = 10.2, yend = 1.15,
           arrow = arrow(length = unit(0.2, "cm")), color = col_demand, linewidth = 0.7) +
  scale_x_continuous(name = "Quantity (billions of pounds)",
                     breaks = seq(7, 15, by = 2),
                     limits = c(6.8, 15.2),
                     expand = expansion(mult = 0.02)) +
  scale_y_continuous(name = "Price ($/lb)",
                     breaks = seq(0.5, 2.0, by = 0.5),
                     limits = c(0.45, 2.05),
                     labels = dollar,
                     expand = expansion(mult = 0.02)) +
  theme_modern_econ()
```

::: {.block}
#### Law of Demand
Higher prices → lower quantity demanded. The demand curve slopes downward.
:::

## Critical Distinction: Movement vs. Shift

```{r}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

# Create shifted demand
demand_shift <- create_demand(intercept = 3.0)

# Movement plot - arrow follows actual demand curve slope
p1 <- ggplot() +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.4) +
  geom_point(aes(x = 10, y = 1.0), size = 4, color = col_demand) +
  geom_point(aes(x = 13, y = 0.55), size = 4, color = col_demand) +
  # Arrow follows demand curve slope: P = 2.5 - 0.15*Q
  geom_segment(aes(x = 10.3, y = 0.955, xend = 12.7, yend = 0.595),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "#E57373", linewidth = 1.2) +
  annotate("text", x = 12.5, y = 1.55, label = "D",
           color = col_demand, size = 5, fontface = "bold") +
  scale_x_continuous(name = "Quantity",
                     limits = c(7, 15), expand = expansion(mult = 0.02)) +
  scale_y_continuous(name = "Price",
                     limits = c(0.4, 1.8), labels = dollar, expand = expansion(mult = 0.02)) +
  ggtitle("Movement Along Curve") +
  theme_modern_econ() +
  theme(plot.title = element_text(size = 11))

# Shift plot - arrow stays within curves
p2 <- ggplot() +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.2, alpha = 0.4) +
  geom_line(data = demand_shift, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.4) +
  # Shorter arrow that doesn't overshoot
  geom_segment(aes(x = 10.5, y = 1.0, xend = 12.5, yend = 1.0),
               arrow = arrow(length = unit(0.25, "cm")),
               color = col_supply, linewidth = 1.2) +
  # Lower D1 and D2 labels
  annotate("text", x = 13, y = 0.85, label = expression(D[1]),
           color = col_demand, size = 5, fontface = "bold", alpha = 0.5) +
  annotate("text", x = 13, y = 1.35, label = expression(D[2]),
           color = col_demand, size = 5, fontface = "bold") +
  scale_x_continuous(name = "Quantity",
                     limits = c(7, 15), expand = expansion(mult = 0.02)) +
  scale_y_continuous(name = "",
                     limits = c(0.4, 1.8), labels = dollar, expand = expansion(mult = 0.02)) +
  ggtitle("Entire Curve Shifts") +
  theme_modern_econ() +
  theme(plot.title = element_text(size = 11),
        axis.title.y = element_blank(),
        axis.text.y = element_blank())

grid.arrange(p1, p2, ncol = 2)
```

\vskip-1em

::: {.block}
**Pro Tip:** If the changing variable is on the Y-axis (price), it's a **movement**. Otherwise, it's a **shift**!
:::

## Common Mistakes: Movement vs. Shift

Students often confuse these concepts. Here's how to avoid errors:

\vskip1em

**❌ WRONG:** "Price increased, so demand shifted"
- **✓ CORRECT:** "Price increased, so quantity demanded decreased" (movement along curve)

\vskip0.5em

**❌ WRONG:** "Income rose, so quantity demanded increased"
- **✓ CORRECT:** "Income rose, so demand increased" (entire curve shifts)

\vskip1em

::: {.block}
#### Quick Test
Ask yourself: "Did **price** change, or did **something else** change?"
- Price changed → **movement** along the curve
- Something else changed → **shift** of the entire curve
:::

## What Shifts Demand? The TRIBE Framework

::: {.columns}
:::: {.column width="48%"}
### Five Demand Shifters

**T** - Tastes and preferences

**R** - Related goods prices

**I** - Income

**B** - Buyers (number of consumers)

**E** - Expectations
::::

:::: {.column width="48%"}
### Remember:

\vskip1em

These factors shift the **entire curve**

\vskip1em

Price changes cause **movements along** the curve

\vskip1em

Direction matters: increase vs. decrease in demand
::::
:::

## Demand Shifter 1: Related Goods

:::: {.columns}
::: {.column width="58%"}
**Substitutes** - Goods that can replace each other

\vskip0.5em

- ↑ Price of Hawaii trips → ↑ Demand for Puerto Rico trips
- ↑ Price of beef → ↑ Demand for chicken
- ↑ Price of iPhone → ↑ Demand for Android phones

\vskip0.5em

::: {.block}
Price of substitute ↑ → Demand ↑
:::

\vskip1em

**Complements** - Goods consumed together

\vskip0.5em

- ↑ Price of cars → ↓ Demand for gasoline
- ↑ Price of printers → ↓ Demand for ink cartridges
- ↑ Price of hot dogs → ↓ Demand for hot dog buns

\vskip0.5em

::: {.block}
Price of complement ↑ → Demand ↓
:::
:::

::: {.column width="40%"}
\begin{center}
\includegraphics[width=\linewidth,height=0.6\textheight,keepaspectratio]{images/smartphone_substitutes.jpg}

{\small \textit{Substitutes: iPhone vs Android}}
\end{center}
:::
::::

## Knowledge Check: Substitutes and Complements

**A new tax raises the price of gasoline. What happens to the demand for electric vehicles?**

\vskip0.5em

a) Demand for electric vehicles increases

b) Demand for electric vehicles decreases

c) Quantity demanded of electric vehicles increases

d) No change in demand for electric vehicles

\vskip2em

\onslide<2->{
::: {.block}
**Answer:** (a)
Gasoline and electric vehicles are substitutes. When the price of gasoline rises, consumers shift toward the substitute → demand for EVs increases (curve shifts right).
:::
}

## Demand Shifter 2: Income

:::: {.columns}
::: {.column width="50%"}
**Normal Goods**

Income ↑ → Demand ↑

\vskip0.5em

Examples:
- Restaurant meals
- New cars
- International travel
- Organic produce

\vskip1em

Most goods are normal goods
:::

::: {.column width="50%"}
**Inferior Goods**

Income ↑ → Demand ↓

\vskip0.5em

Examples:
- Instant ramen
- Used cars (vs. new)
- Generic brands
- Bus rides (vs. owning a car)

\vskip1em

Budget substitutes we trade up from
:::
::::

## Knowledge Check: Inferior Goods

**Which are most likely to be inferior goods?**

\vskip0.5em

a) Store-brand cereal

b) Designer boots

c) Premium deodorant

d) Public bus rides

\vskip2em

\onslide<2->{
::: {.block}
**Answer:** (a) and (d)

Store-brand cereal and bus rides are budget options that people typically replace with premium alternatives (name-brand cereal, personal vehicles) as income rises.
:::
}

## Demand Shifter 3: Tastes and Preferences

Consumer preferences change for many reasons:

\vskip0.5em

**Seasonal patterns**
- Winter → ↑ demand for coats, snow tires
- Summer → ↑ demand for air conditioners, beach gear

\vskip0.5em

**Trends and fads**
- Stanley cups, fidget spinners, Pokémon cards
- These create temporary demand spikes

\vskip0.5em

**New information**
- Health studies linking smoking to cancer → ↓ demand for cigarettes
- Celebrity endorsements → ↑ demand for products

## Demand Shifters 4-5: Expectations and Buyers

**4. Expectations**
- Expect higher future prices → ↑ demand today (buy now!)
- Expect lower future prices → ↓ demand today (wait for sale)
- Expect higher future income → ↑ demand today

\vskip1em

**5. Number of Buyers**
- More consumers → ↑ demand
- Population growth, new markets opening
- Example: More babies → ↑ demand for diapers

# Part II: The Supply Side

## The Supply Curve

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

# Create supply schedule
supply_schedule <- data.frame(
  Price = c(2.00, 1.75, 1.50, 1.25, 1.00, 0.75, 0.50),
  Quantity = c(11.6, 11.5, 11.2, 10.7, 10.0, 9.1, 8.0)
)

ggplot() +
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.5) +
  geom_point(data = supply_schedule, aes(x = Quantity, y = Price),
             color = col_supply, size = 3) +
  annotate("text", x = 13, y = 1.85, label = "Supply",
           color = col_supply, size = 5.5, hjust = 0, fontface = "bold") +
  annotate("text", x = 8.5, y = 1.65,
           label = "As price rises,\nquantity supplied rises",
           color = col_supply, size = 3.5, hjust = 0.5) +
  annotate("segment", x = 9, y = 1.45, xend = 9.5, yend = 1.1,
           arrow = arrow(length = unit(0.25, "cm")), color = col_supply, linewidth = 0.8) +
  scale_x_continuous(name = "Quantity of cotton (billions of pounds)",
                     breaks = seq(7, 15, by = 2),
                     limits = c(7, 15.5),
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Price of cotton (per pound)",
                     breaks = seq(0.5, 2.0, by = 0.25),
                     limits = c(0.45, 2.1),
                     labels = dollar,
                     expand = c(0, 0)) +
  theme_modern_econ()
```

::: {.block}
#### Law of Supply
Higher prices → higher quantity supplied. The supply curve slopes upward.
:::

## Movement vs. Shift: Supply Edition

```{r}
#| fig-width: 10
#| fig-height: 4
#| fig-align: center

# Shifted supply
supply_shift <- create_supply(intercept = 0.0)

# Movement
p3 <- ggplot() +
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.3) +
  geom_point(aes(x = 9, y = 0.85), size = 4, color = col_supply) +
  geom_point(aes(x = 11, y = 1.15), size = 4, color = col_supply) +
  geom_segment(aes(x = 9, y = 0.85, xend = 11, yend = 1.15),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "#E57373", linewidth = 1.2) +
  annotate("text", x = 12.5, y = 1.45, label = "S",
           color = col_supply, size = 5.5, fontface = "bold") +
  annotate("text", x = 10, y = 0.65,
           label = "Movement\nalong curve",
           color = "#E57373", size = 3, hjust = 0.5, fontface = "bold") +
  scale_x_continuous(name = "Quantity",
                     limits = c(7, 15), expand = c(0, 0)) +
  scale_y_continuous(name = "Price",
                     limits = c(0.4, 2.1), labels = dollar, expand = c(0, 0)) +
  ggtitle("Change in Quantity Supplied") +
  theme_modern_econ() +
  theme(plot.title = element_text(size = rel(1)))

# Shift
p4 <- ggplot() +
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.3, alpha = 0.4) +
  geom_line(data = supply_shift, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.3) +
  geom_segment(aes(x = 10, y = 1.0, xend = 13, yend = 1.0),
               arrow = arrow(length = unit(0.3, "cm")),
               color = col_shift, linewidth = 1.2) +
  annotate("text", x = 12.5, y = 1.45, label = expression(S[1]),
           color = col_supply, size = 5.5, fontface = "bold", alpha = 0.4) +
  annotate("text", x = 12.5, y = 1.9, label = expression(S[2]),
           color = col_supply, size = 5.5, fontface = "bold") +
  annotate("text", x = 11.5, y = 0.8,
           label = "Shift of\nentire curve",
           color = col_shift, size = 3, hjust = 0.5, fontface = "bold") +
  scale_x_continuous(name = "Quantity",
                     limits = c(7, 15), expand = c(0, 0)) +
  scale_y_continuous(name = "",
                     limits = c(0.4, 2.1), labels = dollar, expand = c(0, 0)) +
  ggtitle("Change in Supply") +
  theme_modern_econ() +
  theme(plot.title = element_text(size = rel(1)),
        axis.title.y = element_blank())

grid.arrange(p3, p4, ncol = 2)
```

## What Shifts Supply? The TONER Framework

::: {.columns}
:::: {.column width="48%"}
### Five Supply Shifters

**T** - Technology

**O** - Other goods prices

**N** - Number of producers

**E** - Expectations

**R** - Resource costs (inputs)
::::

:::: {.column width="48%"}
### Key Insight:

\vskip1em

Supply shifts when producers' **costs** or **profit opportunities** change

\vskip1em

Better technology, more producers, or lower input costs → ↑ supply

\vskip1em

Higher input costs or fewer producers → ↓ supply
::::
:::

## Supply Shifter 1: Input Prices

:::: {.columns}
::: {.column width="58%"}
**↓ Input Price → ↑ Supply**

Lower production costs mean producers can profitably supply more at each price

\vskip0.5em

Examples:

- ↓ Fertilizer prices → ↑ Cotton supply
- ↓ Steel prices → ↑ Car supply
- ↓ Microchip prices → ↑ Laptop supply

\vskip1em

**↑ Input Price → ↓ Supply**

Higher production costs reduce profitable supply at each price

\vskip0.5em

Examples:

- ↑ Labor costs → ↓ Restaurant supply
- ↑ Fuel costs → ↓ Trucking supply
- ↑ Coffee bean costs → ↓ Latte supply
:::

::: {.column width="40%"}
\begin{center}
\includegraphics[width=\linewidth,height=0.6\textheight,keepaspectratio]{images/factory_production.jpg}

{\small \textit{Input costs affect production}}
\end{center}
:::
::::

## Supply Shifter 2: Related Goods in Production

Producers often can make multiple products (substitutes) or must make them together (complements)

\vskip1em

**Substitutes in Production**

- Farmer's land: corn vs. soybeans
- ↑ Price of soybeans → farmer plants more soybeans → ↓ corn supply

\vskip1em

**Complements in Production**

- Beef and leather (from same cow)
- Pork and lard (from same pig)
- ↑ Demand for beef → ↑ beef production → ↑ leather supply

## Knowledge Check: Production Relationships

**A drought devastates corn crops. Farmers can grow either corn or soybeans on their land. What happens in the soybean market?**

\vskip0.5em

a) Supply of soybeans increases

b) Supply of soybeans decreases

c) Demand for soybeans increases

d) Quantity supplied of soybeans increases

\vskip2em

\onslide<2->{
::: {.block}
**Answer:** (a)
Corn and soybeans are **substitutes in production**. When corn crops fail, farmers switch their land to soybeans → supply of soybeans increases (shifts right).
:::
}

## Supply Shifter 3: Technology

Better technology lowers production costs:

\vskip0.5em

**Examples:**
- Precision agriculture: GPS-guided tractors → less waste, better yields
- Industrial automation: robots → lower labor costs per unit
- AI and machine learning: optimized logistics → reduced shipping costs

\vskip1em

**Key feature:** Technology improvements are usually **permanent**

- Once adopted, they become the new standard
- Competitors must adopt or risk being priced out

\vskip1em

::: {.block}
#### Result
Better technology → ↓ costs → ↑ supply (shift right)
:::

## Supply Shifters 4-5: Expectations and Sellers

**4. Expectations**
- Expect higher future prices → ↓ supply today (save inventory)
- Expect lower future prices → ↑ supply today (sell now!)
- Example: Oil companies withhold production when expecting price increases

\vskip1em

**5. Number of Producers**
- More sellers enter market → ↑ supply
- Sellers exit market → ↓ supply
- Entry/exit depends on profitability and barriers to entry
- Example: Legal marijuana → hundreds of new producers entered market → ↑ supply

## Knowledge Check: Supply Shifters

**Green Mountain Solar knows the government is phasing out solar tax credits in two years. What supply shifter best describes this?**

\vskip0.5em

a) Input prices

b) Prices of related goods

c) Technology

d) Expectations

e) Number of sellers

\vskip2em

\onslide<2->{
::: {.block}
**Answer:** (d) Expectations
Anticipating less favorable conditions in the future affects production decisions today.
:::
}

# Part III: Market Equilibrium

## Finding Equilibrium: Where Supply Meets Demand

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

ggplot() +
  # Supply and demand curves
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.6) +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.6) +
  # Equilibrium point
  geom_point(aes(x = eq_Q, y = eq_P), size = 6, color = col_equilibrium) +
  # Dashed lines to axes
  geom_segment(aes(x = eq_Q, y = 0, xend = eq_Q, yend = eq_P),
               linetype = "dashed", color = "gray50", linewidth = 0.7) +
  geom_segment(aes(x = 0, y = eq_P, xend = eq_Q, yend = eq_P),
               linetype = "dashed", color = "gray50", linewidth = 0.7) +
  # Labels
  annotate("text", x = 13.5, y = 0.85, label = "Demand",
           color = col_demand, size = 6.5, hjust = 0, fontface = "bold") +
  annotate("text", x = 13.5, y = 1.95, label = "Supply",
           color = col_supply, size = 6.5, hjust = 0, fontface = "bold") +
  annotate("text", x = eq_Q + 0.4, y = eq_P + 0.15, label = "E",
           size = 8, fontface = "bold", color = col_equilibrium) +
  annotate("text", x = 6, y = eq_P + 0.05, label = "P*",
           size = 5, fontface = "bold", color = col_equilibrium) +
  annotate("text", x = eq_Q, y = 0.35, label = "Q*",
           size = 5, fontface = "bold", color = col_equilibrium) +
  # Scales
  scale_x_continuous(name = "Quantity of cotton (billions of pounds)",
                     breaks = seq(7, 15, by = 2),
                     limits = c(6.5, 15.5),
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Price of cotton (per pound)",
                     breaks = seq(0.25, 2.0, by = 0.25),
                     limits = c(0.2, 2.1),
                     labels = dollar,
                     expand = c(0, 0)) +
  theme_modern_econ()
```

::: {.block}
**Equilibrium:** Price where quantity demanded = quantity supplied. No shortage, no surplus.
:::

## Surplus: When Price is Too High

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

high_price <- 1.50
Qd_high <- 8.1
Qs_high <- 11.2

ggplot() +
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.5) +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.5) +
  # Equilibrium
  geom_point(aes(x = eq_Q, y = eq_P), size = 4, color = "gray60") +
  # Points at high price
  geom_point(aes(x = Qd_high, y = high_price), size = 4, color = col_demand) +
  geom_point(aes(x = Qs_high, y = high_price), size = 4, color = col_supply) +
  # Surplus bracket
  geom_segment(aes(x = Qd_high, y = high_price + 0.08,
                   xend = Qs_high, yend = high_price + 0.08),
               color = "#E57373", linewidth = 1.5,
               arrow = arrow(ends = "both", length = unit(0.2, "cm"))) +
  annotate("text", x = (Qd_high + Qs_high)/2, y = high_price + 0.22,
           label = "SURPLUS",
           color = "#E57373", size = 5, fontface = "bold") +
  # Horizontal line at high price
  geom_segment(aes(x = 0, y = high_price, xend = Qs_high, yend = high_price),
               linetype = "dashed", color = "gray50", linewidth = 0.6) +
  # Labels
  annotate("text", x = Qd_high, y = high_price - 0.15,
           label = "Qd", color = col_demand, size = 4, fontface = "bold") +
  annotate("text", x = Qs_high, y = high_price - 0.15,
           label = "Qs", color = col_supply, size = 4, fontface = "bold") +
  annotate("text", x = 13.5, y = 0.85, label = "D",
           color = col_demand, size = 6, hjust = 0, fontface = "bold") +
  annotate("text", x = 13.5, y = 1.95, label = "S",
           color = col_supply, size = 6, hjust = 0, fontface = "bold") +
  annotate("text", x = eq_Q + 0.3, y = eq_P, label = "E",
           size = 6, fontface = "bold", color = "gray60") +
  scale_x_continuous(name = "Quantity",
                     breaks = c(Qd_high, eq_Q, Qs_high),
                     labels = c("8.1", "10.0", "11.2"),
                     limits = c(7, 15.5),
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Price",
                     breaks = c(eq_P, high_price),
                     labels = c("$1.00", "$1.50"),
                     limits = c(0.45, 2.0),
                     expand = c(0, 0)) +
  theme_modern_econ()
```

::: {.block}
**Surplus** → Sellers lower prices to clear excess inventory → moves toward equilibrium
:::

## Shortage: When Price is Too Low

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

low_price <- 0.75
Qd_low <- 11.5
Qs_low <- 9.1

ggplot() +
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.5) +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.5) +
  # Equilibrium
  geom_point(aes(x = eq_Q, y = eq_P), size = 4, color = "gray60") +
  # Points at low price
  geom_point(aes(x = Qs_low, y = low_price), size = 4, color = col_supply) +
  geom_point(aes(x = Qd_low, y = low_price), size = 4, color = col_demand) +
  # Shortage bracket
  geom_segment(aes(x = Qs_low, y = low_price - 0.08,
                   xend = Qd_low, yend = low_price - 0.08),
               color = "#E57373", linewidth = 1.5,
               arrow = arrow(ends = "both", length = unit(0.2, "cm"))) +
  annotate("text", x = (Qs_low + Qd_low)/2, y = low_price - 0.20,
           label = "SHORTAGE",
           color = "#E57373", size = 5, fontface = "bold") +
  # Horizontal line at low price
  geom_segment(aes(x = 0, y = low_price, xend = Qd_low, yend = low_price),
               linetype = "dashed", color = "gray50", linewidth = 0.6) +
  # Labels
  annotate("text", x = Qs_low, y = low_price + 0.12,
           label = "Qs", color = col_supply, size = 4, fontface = "bold") +
  annotate("text", x = Qd_low, y = low_price + 0.12,
           label = "Qd", color = col_demand, size = 4, fontface = "bold") +
  annotate("text", x = 13.5, y = 0.85, label = "D",
           color = col_demand, size = 6, hjust = 0, fontface = "bold") +
  annotate("text", x = 13.5, y = 1.95, label = "S",
           color = col_supply, size = 6, hjust = 0, fontface = "bold") +
  annotate("text", x = eq_Q + 0.3, y = eq_P, label = "E",
           size = 6, fontface = "bold", color = "gray60") +
  scale_x_continuous(name = "Quantity",
                     breaks = c(Qs_low, eq_Q, Qd_low),
                     labels = c("9.1", "10.0", "11.5"),
                     limits = c(7, 15.5),
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Price",
                     breaks = c(low_price, eq_P),
                     labels = c("$0.75", "$1.00"),
                     limits = c(0.45, 2.0),
                     expand = c(0, 0)) +
  theme_modern_econ()
```

::: {.block}
**Shortage** → Buyers bid prices up to secure scarce goods → moves toward equilibrium
:::

## Knowledge Check: Shortage or Surplus?

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

# Generic supply and demand
gen_supply <- create_supply(intercept = 1, slope = 0.3, Q_range = c(4, 20))
gen_demand <- create_demand(intercept = 7, slope = -0.3, Q_range = c(4, 20))

ggplot() +
  geom_line(data = gen_supply, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.4) +
  geom_line(data = gen_demand, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.4) +
  geom_point(aes(x = 10, y = 4), size = 4, color = col_equilibrium) +
  annotate("text", x = 16, y = 2.4, label = "D",
           color = col_demand, size = 5.5, fontface = "bold") +
  annotate("text", x = 16, y = 7.0, label = "S",
           color = col_supply, size = 5.5, fontface = "bold") +
  annotate("text", x = 10.5, y = 4.4, label = "E",
           size = 6, fontface = "bold", color = col_equilibrium) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "#E57373", linewidth = 0.8) +
  annotate("text", x = 18, y = 5.3, label = "Price = $5",
           color = "#E57373", size = 4, fontface = "bold") +
  scale_x_continuous(name = "Quantity",
                     limits = c(4, 20), expand = c(0, 0)) +
  scale_y_continuous(name = "Price",
                     limits = c(1, 8), labels = dollar, expand = c(0, 0)) +
  theme_modern_econ()
```

**At \$5, is there a shortage or surplus?**

\onslide<2->{
::: {.block}
**Answer: Surplus.** At \$5 (above equilibrium), Qs > Qd → excess supply.
:::
}

# Part IV: Putting It All Together

## When Demand Increases

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

demand_inc <- create_demand(intercept = 3.0)
eq_Q2 <- 12.5
eq_P2 <- 1.375

ggplot() +
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.5) +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.3, alpha = 0.5, linetype = "dashed") +
  geom_line(data = demand_inc, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.5) +
  geom_point(aes(x = eq_Q, y = eq_P), size = 5, color = "gray60") +
  geom_point(aes(x = eq_Q2, y = eq_P2), size = 5, color = col_equilibrium) +
  geom_segment(aes(x = 10, y = 1.0, xend = 12, yend = 1.0),
               arrow = arrow(length = unit(0.3, "cm")),
               color = col_shift, linewidth = 1.3) +
  annotate("text", x = 13, y = 0.85, label = expression(D[1]),
           color = col_demand, size = 5, alpha = 0.5) +
  annotate("text", x = 13, y = 1.4, label = expression(D[2]),
           color = col_demand, size = 5, fontface = "bold") +
  annotate("text", x = 13, y = 1.85, label = "S",
           color = col_supply, size = 5, fontface = "bold") +
  scale_x_continuous(name = "Quantity", limits = c(7, 15), expand = c(0, 0)) +
  scale_y_continuous(name = "Price", limits = c(0.45, 2.0), labels = dollar, expand = c(0, 0)) +
  theme_modern_econ()
```

::: {.block}
**Result:** Price ↑ and Quantity ↑ **(Both certain)**
:::

\vskip-0.5em
{\small \textit{Animation: See `images/demand\_increase.gif` for dynamic visualization}}

## When Supply Decreases

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

supply_dec <- create_supply(intercept = -0.1, slope = 0.15)
eq_Q3 <- 8.67
eq_P3 <- 1.20

ggplot() +
  geom_line(data = supply_std, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.3, alpha = 0.5, linetype = "dashed") +
  geom_line(data = supply_dec, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.5) +
  geom_line(data = demand_std, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.5) +
  geom_point(aes(x = eq_Q, y = eq_P), size = 5, color = "gray60") +
  geom_point(aes(x = eq_Q3, y = eq_P3), size = 5, color = col_equilibrium) +
  geom_segment(aes(x = 10, y = 1.0, xend = 8.5, yend = 1.0),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "#E57373", linewidth = 1.3) +
  annotate("text", x = 13, y = 1.45, label = expression(S[1]),
           color = col_supply, size = 5, alpha = 0.5) +
  annotate("text", x = 13, y = 1.85, label = expression(S[2]),
           color = col_supply, size = 5, fontface = "bold") +
  annotate("text", x = 13, y = 0.85, label = "D",
           color = col_demand, size = 5, fontface = "bold") +
  scale_x_continuous(name = "Quantity", limits = c(7, 15), expand = c(0, 0)) +
  scale_y_continuous(name = "Price", limits = c(0.45, 2.0), labels = dollar, expand = c(0, 0)) +
  theme_modern_econ()
```

::: {.block}
**Result:** Price ↑ and Quantity ↓ **(Both certain)**
:::

\vskip-0.5em
{\small \textit{Animation: See `images/supply\_decrease.gif` for dynamic visualization}}

## Practice: Predicting Market Changes

**If the cost of wood falls, what happens to equilibrium price and quantity in the violin market?**

\vskip0.5em

Think about:
1. Which curve shifts? (Supply or demand?)
2. Which direction? (Left or right?)
3. What happens to P* and Q*?

\vskip2em

\onslide<2->{
::: {.block}
**Answer:**
Wood is an input → ↓ input prices → ↑ supply (shift right) → ↓ P* and ↑ Q*
:::
}

## Simultaneous Shifts: Handle with Care

**Example: iPhone launch**

When Apple releases a new iPhone:
1. Consumer excitement → **demand increases** (curve shifts right)
2. Apple ramps up production → **supply increases** (curve shifts right)

\vskip1em

**What happens to equilibrium?**

:::: {.columns}
::: {.column width="48%"}
Both shifts push **quantity** in the same direction:
- ↑ Demand → Q ↑
- ↑ Supply → Q ↑
- **Result: Q ↑ (certain)**
:::

::: {.column width="48%"}
But shifts push **price** in opposite directions:
- ↑ Demand → P ↑
- ↑ Supply → P ↓
- **Result: P ? (ambiguous)**
:::
::::

\vskip0.5em

::: {.block}
#### General Rule
When both curves shift, if the effects on P or Q work in the **same direction** → certain change. If **opposite directions** → ambiguous (depends on relative magnitudes).
:::

\vskip0.5em

**Interactive Demo:** Open `interactive_simultaneous_shifts_v2.html` - use sliders to shift curves and see effects decomposed!

# Real-World Application: The Vanilla Crisis

## Remember This?

**How much does a 4oz bottle of pure vanilla extract cost?**

a) \$4

b) \$7

c) \$10

d) \$18

e) \$20

## The Vanilla Market: 2017

::: {.columns}
:::: {.column width="35%"}
\begin{center}
\includegraphics[width=\linewidth,height=0.65\textheight,keepaspectratio]{images/vanilla_beans.jpg}

{\small \textit{Madagascar vanilla beans}}
\end{center}
::::

:::: {.column width="65%"}
**What happened:**

- Vanilla prices: \$20/kg (2010) → \$500/kg (2017)
- **25× increase in 7 years**

\vskip0.5em

**Why:**

1. Cyclone damaged 30% of Madagascar crop (supply ↓)
2. "All-natural" food trend (demand ↑)
3. Major brands (Nestlé, Hershey's) switched from synthetic to natural vanilla (demand ↑↑)

\vskip0.5em
```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

# Vanilla price data (simplified)
vanilla_data <- data.frame(
  Year = c(2010, 2012, 2014, 2015, 2016, 2017),
  Price = c(20, 25, 30, 50, 200, 500)
)

ggplot(vanilla_data, aes(x = Year, y = Price)) +
  geom_line(color = col_supply, linewidth = 1.3) +
  geom_point(color = col_supply, size = 3) +
  scale_y_continuous(name = "Price ($/kg)",
                     limits = c(0, 600),
                     labels = dollar,
                     expand = c(0, 0)) +
  scale_x_continuous(name = "",
                     breaks = c(2010, 2012, 2014, 2016, 2017)) +
  ggtitle("Vanilla Bean Prices") +
  theme_modern_econ()
```
::::
:::

## Vanilla: Supply and Demand Analysis

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

# Stylized vanilla market
vanilla_D1 <- create_demand(intercept = 80, slope = -4, Q_range = c(5, 20))
vanilla_D2 <- create_demand(intercept = 120, slope = -4, Q_range = c(5, 25))
vanilla_S1 <- create_supply(intercept = 10, slope = 3, Q_range = c(5, 20))
vanilla_S2 <- create_supply(intercept = 20, slope = 3, Q_range = c(5, 20))

# Original eq: S1 and D1 intersect around Q=10, P=40
# New eq: S2 and D2 intersect around Q=13, P=70

ggplot() +
  # Original curves (faded)
  geom_line(data = vanilla_S1, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.2, alpha = 0.4, linetype = "dashed") +
  geom_line(data = vanilla_D1, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.2, alpha = 0.4, linetype = "dashed") +
  # New curves
  geom_line(data = vanilla_S2, aes(x = Q, y = P),
            color = col_supply, linewidth = 1.5) +
  geom_line(data = vanilla_D2, aes(x = Q, y = P),
            color = col_demand, linewidth = 1.5) +
  # Original equilibrium
  geom_point(aes(x = 10, y = 40), size = 4, color = "gray60") +
  # New equilibrium
  geom_point(aes(x = 13.33, y = 70), size = 5, color = col_equilibrium) +
  # Arrows
  geom_segment(aes(x = 10, y = 40, xend = 8, yend = 40),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "#E57373", linewidth = 1) +
  geom_segment(aes(x = 10, y = 40, xend = 16, yend = 40),
               arrow = arrow(length = unit(0.25, "cm")),
               color = col_shift, linewidth = 1) +
  annotate("text", x = 6.5, y = 37,
           label = "Supply ↓\n(cyclone)", color = "#E57373", size = 3.5, fontface = "bold") +
  annotate("text", x = 17.5, y = 37,
           label = "Demand ↑\n(all-natural trend)", color = col_shift, size = 3.5, fontface = "bold") +
  # Labels
  annotate("text", x = 18.5, y = 20, label = expression(D[1]),
           color = col_demand, size = 5, alpha = 0.4) +
  annotate("text", x = 22, y = 40, label = expression(D[2]),
           color = col_demand, size = 5, fontface = "bold") +
  annotate("text", x = 18.5, y = 65, label = expression(S[1]),
           color = col_supply, size = 5, alpha = 0.4) +
  annotate("text", x = 18.5, y = 75, label = expression(S[2]),
           color = col_supply, size = 5, fontface = "bold") +
  annotate("text", x = 9.5, y = 35, label = "2010",
           size = 4, color = "gray60") +
  annotate("text", x = 14, y = 74, label = "2017",
           size = 4.5, fontface = "bold", color = col_equilibrium) +
  scale_x_continuous(name = "Quantity",
                     limits = c(5, 25), expand = c(0, 0)) +
  scale_y_continuous(name = "Price ($/kg)",
                     limits = c(0, 140), labels = dollar, expand = c(0, 0)) +
  ggtitle("The Vanilla Crisis: Simultaneous Shifts") +
  theme_modern_econ()
```

::: {.block}
Supply ↓ **and** Demand ↑ → Price ↑↑ (certain), Quantity ? (ambiguous, but increased)
:::

## The Market Response

By 2019-2020, vanilla prices began falling:

\vskip1em

**Why?**

1. **High prices incentivized production** → farmers planted more vanilla → ↑ supply

2. **Substitution** → some manufacturers switched back to synthetic vanillin → ↓ demand

3. **Recovery** → Madagascar crop recovered → ↑ supply

\vskip1.5em

::: {.block}
#### Classic Market Dynamics
High prices signal scarcity → encourage production and substitution → prices adjust back toward equilibrium
:::

# Conclusion

## Key Takeaways

1. **Demand curves slope down**, supply curves slope up

2. **Movements** happen along curves when price changes; **shifts** happen when non-price factors change

3. **TRIBE** shifts demand: Tastes, Related goods, Income, Buyers, Expectations

4. **TONER** shifts supply: Technology, Other goods, Number of producers, Expectations, Resource costs

5. **Equilibrium** balances Qd and Qs; surpluses and shortages push price toward equilibrium

6. **Price and quantity changes** depend on which curve shifts and in what direction

## Final Knowledge Check

**Garden gnomes regain popularity. What happens to equilibrium?**

a) P ↑, Q ↑

b) P ↑, Q ↓

c) P ↓, Q ↑

d) P ↓, Q ↓

\vskip2em

\onslide<2->{
::: {.block}
**Answer: (a)** Tastes → demand shifts right → P ↑ and Q ↑

This is the type of reasoning you'll use throughout the course!
:::
}

## Next Time

**Chapter 6: Elasticity**

- How responsive is quantity to price changes?
- Why do some price changes barely affect quantity while others cause huge swings?
- What determines whether sellers or buyers bear the burden of a tax?

\vskip1.5em

::: {.block}
### Before Next Class
- Review TRIBE and TONER
- Practice: For each news story you see this week, ask: "Which curve shifted? Which direction?"
:::

#

\begin{center}
{\Huge\textbf{Thank You}}

\vskip2em

Questions?

\vskip3em

{\large ECON 1450: Principles of Microeconomics}

University of Vermont

\end{center}
