---
title: "Supply and Demand"
subtitle: "ECON 1450: Principles of Microeconomics"
author: "University of Vermont"
date: "Fall 2026"
institute: "Department of Economics"
format:
  beamer:
    theme: ElegantProfessional
    aspectratio: 169
    navigation: horizontal
    section-titles: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(dplyr)
library(scales)

theme_econ <- function(base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = rel(1.2)),
      plot.background = element_rect(fill = "#FFFFF0", color = NA),
      panel.background = element_rect(fill = "#FFFFF0", color = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#D3D3D3", linewidth = 0.3),
      axis.line = element_line(color = "#2F2F2F", linewidth = 0.7),
      axis.text = element_text(size = rel(1), color = "#2F2F2F"),
      axis.title = element_text(size = rel(1.1), face = "bold"),
      axis.ticks = element_line(color = "#2F2F2F", linewidth = 0.5),
      legend.position = "none"
    )
}

demand_curve <- data.frame(Q = seq(7, 15, by = 0.05)) %>%
  mutate(P = 2.5 - 0.15 * Q)
supply_curve <- data.frame(Q = seq(7, 15, by = 0.05)) %>%
  mutate(P = -0.5 + 0.15 * Q)
eq_Q <- 10
eq_P <- 1.0
```

# THEME 4: Elegant Professional

##  {.plain}

\begin{center}
\Large\textbf{Theme Sampler}

\vskip1em

\normalsize
This presentation demonstrates the \textbf{Elegant Professional} theme

\vskip0.5em

Sophisticated palette • Balanced typography • Gold accents

\vskip1em

{\color{structure}Ideal for: professional settings, executive presentations}
\end{center}

## Learning Objectives

By the end of this lecture, you will be able to:

- Define **competitive markets** and the supply-demand model
- Distinguish between **movements along** curves and **shifts** of curves
- Identify the five demand shifters and five supply shifters
- Find market **equilibrium** and analyze shortages and surpluses
- Predict price and quantity changes from market shocks

## Two-Column Layout Example

:::: {.columns}
::: {.column width="48%"}
### Demand Shifters

**TRIBE** mnemonic:

- **T**astes
- **R**elated goods prices
- **I**ncome
- **B**uyers (# of consumers)
- **E**xpectations
:::

::: {.column width="48%"}
### Supply Shifters

**TONER** mnemonic:

- **T**echnology
- **O**ther goods prices
- **N**umber of producers
- **E**xpectations
- **R**esource costs
:::
::::

\vskip1em

::: {.block}
#### Pro Tip
If the changing variable is on the Y-axis (price), it's a **movement**. Otherwise, it's a **shift**!
:::

## Supply and Demand: Market Equilibrium

```{r}
#| fig-width: 9
#| fig-height: 5.5
#| fig-align: center

ggplot() +
  geom_line(data = supply_curve, aes(x = Q, y = P),
            color = "#6A5ACD", linewidth = 1.6) +
  geom_line(data = demand_curve, aes(x = Q, y = P),
            color = "#808076", linewidth = 1.6) +
  geom_point(aes(x = eq_Q, y = eq_P), size = 6, color = "#D4AF37") +
  geom_segment(aes(x = eq_Q, y = 0, xend = eq_Q, yend = eq_P),
               linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_segment(aes(x = 0, y = eq_P, xend = eq_Q, yend = eq_P),
               linetype = "dashed", color = "gray50", linewidth = 0.6) +
  annotate("text", x = 13.5, y = 0.9, label = "Demand",
           color = "#808076", size = 6, hjust = 0, fontface = "italic") +
  annotate("text", x = 13.5, y = 1.8, label = "Supply",
           color = "#6A5ACD", size = 6, hjust = 0, fontface = "italic") +
  annotate("text", x = eq_Q + 0.3, y = eq_P + 0.15, label = "E",
           size = 7, fontface = "bold", color = "#D4AF37") +
  scale_x_continuous(name = "Quantity (billions of units)",
                     breaks = seq(7, 15, by = 2),
                     limits = c(7, 15.5),
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Price ($ per unit)",
                     breaks = seq(0.5, 2.5, by = 0.5),
                     limits = c(0.45, 2.6),
                     labels = dollar,
                     expand = c(0, 0)) +
  theme_econ()
```

## Knowledge Check

**Question:** If garden gnomes regain popularity, what happens to equilibrium price and quantity?

\vskip1em

a) Price ↑, Quantity ↑

b) Price ↑, Quantity ↓

c) Price ↓, Quantity ↑

d) Price ↓, Quantity ↓

\vskip2em

. . .

::: {.block}
#### Answer: (a) Price ↑, Quantity ↑

**Reasoning:** Tastes shift demand right → higher willingness to pay at every quantity → new equilibrium at higher price and quantity
:::

## Summary: Elegant Professional Theme

**Design Features:**

- Sophisticated slate blue and warm gray palette
- Gold accent lines for visual interest
- Balanced serif/sans-serif font combination
- Subtle ivory background
- Professional footer with complete navigation

**Best For:**

- Professional/business settings
- Executive presentations
- Policy briefings

**Graph Style:** Refined with muted colors, gold equilibrium point
