---
title: "Supply and Demand"
subtitle: "ECON 1450: Principles of Microeconomics"
author: "Theme Sampler - 5 Design Options"
date: today
format:
  beamer:
    theme: default
    colortheme: default
    aspectratio: 169
    navigation: horizontal
    section-titles: false
    incremental: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(dplyr)
library(scales)

# Custom theme for economics graphs
theme_econ <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
      axis.line = element_line(color = "black", linewidth = 0.5),
      axis.text = element_text(size = rel(0.9)),
      axis.title = element_text(size = rel(1)),
      legend.position = "none"
    )
}

# Supply and demand data
create_supply_demand <- function() {
  # Demand curve data
  demand <- data.frame(
    Q = seq(7, 15, by = 0.1)
  ) %>%
    mutate(P = 2.5 - 0.15 * Q)

  # Supply curve data
  supply <- data.frame(
    Q = seq(7, 15, by = 0.1)
  ) %>%
    mutate(P = -0.5 + 0.15 * Q)

  list(demand = demand, supply = supply)
}

curves <- create_supply_demand()
```

# Theme Overview

## Five Design Options

This sampler demonstrates five distinct Beamer themes for ECON 1450:

1. **Classic Academic** - Traditional, serif-based
2. **Modern Minimal** - Clean, contemporary
3. **Bold Visual** - Eye-catching, geometric
4. **Elegant Professional** - Sophisticated business style
5. **Data-First** - Optimized for graphs and charts

Each theme is shown with identical content for comparison.

---

:::{.callout-note}
**How to use this sampler:** Each theme shows the same 5 slide types: title, content, two-column, graph, and knowledge check. Choose your favorite or mix elements.
:::

# THEME 1: Classic Academic

## Learning Objectives

By the end of this lecture, you will be able to:

- Define competitive markets and the supply-demand model
- Distinguish between movements along curves and shifts of curves
- Identify the five demand shifters and five supply shifters
- Find market equilibrium and analyze shortages and surpluses
- Predict price and quantity changes from market shocks

## Two-Column Layout

:::: {.columns}
::: {.column width="50%"}
### Demand Shifters (TRIBE)

- **T**astes
- **R**elated goods prices
- **I**ncome
- **B**uyers (# of consumers)
- **E**xpectations
:::

::: {.column width="50%"}
### Supply Shifters (TONER)

- **T**echnology
- **O**ther goods prices
- **N**umber of producers
- **E**xpectations
- **R**esource costs (inputs)
:::
::::

## Supply and Demand Curves

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

ggplot() +
  geom_line(data = curves$demand, aes(x = Q, y = P),
            color = "#2E5090", linewidth = 1.2) +
  geom_line(data = curves$supply, aes(x = Q, y = P),
            color = "#C41E3A", linewidth = 1.2) +
  geom_point(aes(x = 10, y = 1.0), size = 4, color = "black") +
  annotate("text", x = 13, y = 1.0, label = "Demand",
           color = "#2E5090", size = 5, hjust = 0) +
  annotate("text", x = 13, y = 1.75, label = "Supply",
           color = "#C41E3A", size = 5, hjust = 0) +
  annotate("text", x = 10, y = 0.85, label = "Equilibrium",
           size = 4, hjust = 0.5) +
  scale_x_continuous(name = "Quantity",
                     breaks = seq(7, 15, by = 2),
                     limits = c(7, 15)) +
  scale_y_continuous(name = "Price",
                     breaks = seq(0.5, 2.5, by = 0.5),
                     limits = c(0.5, 2.5),
                     labels = dollar) +
  theme_econ()
```

## Knowledge Check

**If garden gnomes regain popularity, what happens to equilibrium price and quantity?**

a) Price ↑, Quantity ↑
b) Price ↑, Quantity ↓
c) Price ↓, Quantity ↑
d) Price ↓, Quantity ↓

:::{.fragment}
**Answer: a)** Tastes shift demand right → higher price and quantity
:::

---

:::{.callout-tip}
## Theme 1 Summary
**Classic Academic** uses traditional colors, serif fonts, and clean layouts. Best for: formal presentations, academic conferences.
:::

# Additional Sampler Content

## Theme Comparison Notes

The remaining themes (Modern Minimal, Bold Visual, Elegant Professional, Data-First) would be shown in separate rendered PDFs, each with:

- Distinct color palettes
- Different font choices
- Varied layout approaches
- Same content for direct comparison

## Next Steps

1. Review all 5 theme PDFs
2. Select favorite or identify preferred elements
3. I'll create custom theme file based on your selection
4. Build full Chapter 3 deck with chosen design

---

**Note:** To generate all 5 themes, I'll create separate .qmd files (one per theme) that compile to individual PDFs. This allows side-by-side comparison.
